Optimal updating magnitude in adaptive flat-distribution-sampling simulations

### 2017-06-21 doc/invt.tex
add update.sh for collecting LJ data

\subsubsection{\label{sec:optscheme}
Bandpass updating schemes}

@@

In addition to the application to homogeneous updating schemes
(Appendix \ref{sec:homo_bandpass}),
bandpass updating schemes can also be applied to
a continuous variable, $z$,\footnote{
  In the transition to a continuous variable,
  we will map
  $\rho_i$ to $\rho(z) \delta z$,
  $\phi_{ki}$ to $\phi_k(z) \delta z$,
  $w_{ij}$ to $w(z, z') \delta z$, etc.}
with the eigenmodes derived from orthogonal polynomials, $R_k(z)$,
satisfying
\begin{align*}
  \int R_k(z) R_{k'}(z) \rho(z) \, d z = \delta_{k, k'}
  .
\end{align*}
%
With $\phi_k(z) = R_k(z) \, \rho(z)$,
we can readily satisfy Eqs. \eqref{eq:eig_orthonormal_cols}-\eqref{eq:ortho0},
and construct the updating matrix by
Eqs. \eqref{eq:w_from_phi} and \eqref{eq:lambda_bandpass} as
\begin{align*}
  w(z, z')
  =
  \sum_{k=1}^{K-1} \frac{ \phi_k(z) \, \phi_k(z') } { \rho(z) }
  = \sum_{k=1}^{K-1} R_k(z) \, R_k(z') \, \rho(z'),
\end{align*}
%
where we have dropped the $k=0$ mode
that corresponds to the uniform shift of the bias potential.
%
This results in a polynomial bias potential
that can be written as a superposition
$$
u(z, t) = \sum_{k=1}^{K-1} c_k(t) \, R_k(z),
$$
with the coefficients updated as, according to Eq. \eqref{eq:mbin_update},
$$
c_k(t+1) = c_k(t) + \alpha(t) \, R_k\bigl( z(t) \bigr).
$$

For example,
to sample a flat distribution, $\rho(z)$,
over $[z_c - \Delta z/2, z_c + \Delta z/2]$,
we can use the Legendre polynomials\cite{arfken},
and with $K = 2$,
the bias potential contains only a linear mode for
$R_1(z) = 2 \sqrt{3} (z - z_c)/\Delta z$,
which recovers the algorithm by
Langfeld, Lucini and Rago\cite{langfeld2012}.
%
Applied to the energy space, $z = E$,
$c_1(t)$ represents the scaled local or
statistical temperature\cite{kim2006, *kim2007},
and the inverse-time schedule
was recommended for its update\cite{pellegrini2014}.

Alternatively,
to sample a Gaussian distribution
centered at $z_c$ with width $\sigma_z$
we can use the Hermite polynomials\cite{arfken},
and with $K=3$,
we have a quadratic bias potential
from a linear combination of
$R_1(z) = (z - z_c)/\sigma_z$
and
$R_2(z) = (z - z_c)^2 /\left(\sqrt 2 \, \sigma_z^2\right)$.
%
This bias potential was commonly empolyed
in umbrella sampling (see e.g. Ref. \cite{zhu2012})
and the method of Gaussian ensemble\cite{hetherington1987,
*challa1988, *costeniuc2006, neuhaus2006, *neuhaus2007},
which is helpful in overcoming barriers
in first-order phase transitions.
%
While the asymptotic updating magntiude
would be given by the inverse-time formula,
we can employ the WL strategy early in simulation
to control the updating magnitude in stages
with the histogram fluctuation redefined as
$$
\max\left\{
  \left|
    \frac{ \overline z - z_c } { \sigma_z }
  \right|
  ,
  \left|
    \frac{ \overline{ (z - z_c)^2 } - \sigma_z^2 }
    { \sqrt 2 \, \sigma_z^2 }
  \right|
  \right\}
  ,
$$
where the overline denotes the time average.

With only two nontrivial modes,
the above strategy usually only works
for a Gaussian with a narrow width.
%
We can, however, use it as a building block
to construct an expanded ensemble with
multiple Gaussians centered at different places.
%
Such a construction can be realized via
either the replica exchange
(parallel tempering)\cite{swendsen1986,
*geyer1991, *hukushima1996, *hansmann1997}
or the simulated tempering\cite{marinari1992,
  *lyubartsev1992, li2007,
  park2007, *nguyen2013, *zhang2015st}
framework.
While the mutual exchange probabiliy
can be readily derived in the former
framework\cite{neuhaus2006, *neuhaus2007},
the transition probability in the latter case is
requires estimation of the partition function
of the Gaussian ensemble
$Z \equiv \int e^{-\beta \, E(\mathbf x)-u(z(\mathbf x),t)} \, d\mathbf x$,
where $\mathbf x$ denotes the molecular coordinates,
$E(\mathbf x)$ is the unbiased energy function,
$\beta$ is the inverse temperature,
and the collective variable, $z$,
has also been written as a function of $\mathbf x$.
%
The transition probabiliy from
the $i$th Gaussian to the $j$th is given by
is given by
\begin{align*}
A(i\to j) =
\min\left\{1,
  \frac{Z^{(i)}}{Z^{j}}
  \exp\left[u^{(i)}\bigl( z(t) \bigr)
   - u^{(j)}\bigl( z(t) \bigr) \right]
  \right\}
  ,
\end{align*}
where $z(t)$ is the current value of $z$.
%
For a neighboring pair, $j = i\pm 1$,
we can estimate the ratio $Z^{(i)}/Z^{(j)}$
from the following approximate formula
%
\begin{align*}
  \frac{Z_i}{Z_j}
  =
  \frac{\sigma_i}{\sigma_j}
  \exp\left[
    u^{(j)}\bigl( z^{(i,j)} \bigr)
   -u^{(i)}\bigl( z^{(i,j)} \bigr)
  \right]
  ,
\end{align*}
%
where
$z^{(i,j)} \equiv
\bigl( z_c^{(i)}/\sigma_z^{(i)} + z_c^{(j)}/\sigma_z^{(j)} \bigr) /
\bigl(         1/\sigma_z^{(i)} +         1/\sigma_z^{(j)} \bigr)$.

@@

%\subsection{\label{sec:lj}Lennard-Jones}

@@

In Fig. \ref{fig:lj_alpha},
we show the optimal schedules in a few cases.
%
%Generally, the optimal schedule
%for the $\varrho^* = 0.8$ case
%was closer to the inverse-time schedule
%than that for the $\varrho^* = 0.1$ case.
%%
%This was probably because in the former case
%the first mode dominates the total error,
%$\Gamma_1 \approx \Gamma$,
%and the schedule was roughly optimized for this mode,
%resulting in
%$\alpha(t) \approx 1/[\lambda_1 (t + t_0)] \approx 1/(t+t_0)$
%as $\lambda_1 \approx 1$
%[cf. Eq. \eqref{eq:lambda_Gaussian}].
%
Since we have chosen a long equilibration period
for the preliminary run under constant updating magnitude,
the starting magnitude was around
the ideal value of $a_0/2 = 5\times10^{-5}$,
as given by Eq. \eqref{eq:half_alpha0}.
%
But if the Gaussian were too wide,
e.g. if $\sigma = 0.6$,
the eigenvalues $\lambda_k$ of some modes
became too small to damp out the initial bias,
resulting in much larger initial updating magnitude, $\alpha(0)$
[cf. Eq. \eqref{eq:opt_alpha0}].
%
In this case, the schedule predicted by our method
would be inaccurate because the resulting deviation in the bias potential
was no longer small,
and one may have to either decrease the width
or to elongate the preliminary simulation under constant magnitude.

@@

@@

The relation between error and histogram fluctuation
supports the WL criterion for
checking convergence at stage transitions.
%
For bandpass updating scheme,
we may borrow this criterion
in initial stages, with a modified definition
for the histogram fluctuation
$F^2_{<K}(T) \equiv \sum_{k=1}^{K-1} \tilde F_k^2(T),$
since the histogram flattening
is only partially applied to modes less $K$.
%
One can alternatively define the histogram fluctuation
from the largest mode
$$
\max_{k=1, \dots, K-1} \left| \, \tilde F_k(T) \, \right|
.
$$
For a local sampling process,
we expect that the two definition would behave similarly.
%
\note{
Note that the expected histogram flatness
was previously used to define the error\cite{zhou2005, zhou2008},
and the above demonstration shows the definition
is compatible with ours, Eq. \eqref{eq:error_def}, in the long-time limit.
An interesting application\cite{zhou2008},
of the above result is
to set the updating magnitude by $F/\Gamma$
for some empirical value [$\Gamma \approx 10$ there, see Eq. (8)]
as a substitute of the inverse-time schedule.
}

>>>>

\subsubsection{\label{sec:bandpass}
Bandpass updating schemes}

@@

In addition to the application to homogeneous updating schemes
(Appendix \ref{sec:homo_bandpass}),
bandpass updating schemes can also be applied to
a continuous variable, $z$,\footnote{
  In the transition to a continuous variable,
  we will map
  $\rho_i$ to $\rho(z) \delta z$,
  $\phi_{ki}$ to $\phi_k(z) \delta z$,
  $w_{ij}$ to $w(z, z') \delta z$, etc.}
with the eigenmodes derived from orthogonal polynomials, $R_k(z)$,
that satisfy
\begin{align*}
  \int R_k(z) \, R_{k'}(z) \, \rho(z) \, d z = \delta_{k, k'}
  .
\end{align*}
%
With $\phi_k(z) = R_k(z) \, \rho(z)$,
we can readily satisfy Eqs. \eqref{eq:eig_orthonormal_cols}-\eqref{eq:ortho0},
and construct the updating matrix by
Eqs. \eqref{eq:w_from_phi} and \eqref{eq:lambda_bandpass} as
\begin{align*}
  w(z, z')
  =
  \sum_{k=1}^{K-1} \frac{ \phi_k(z) \, \phi_k(z') } { \rho(z) }
  = \sum_{k=1}^{K-1} R_k(z) \, R_k(z') \, \rho(z'),
\end{align*}
%
where we have dropped the $k=0$ mode
that corresponds to the uniform shift of the bias potential.
%
This results in a polynomial bias potential
that can be written as a superposition
$$
u(z, t) = \sum_{k=1}^{K-1} c_k(t) \, R_k(z),
$$
with the coefficients updated as, according to Eq. \eqref{eq:mbin_update},
$$
c_k(t+1) = c_k(t) + \alpha(t) \, R_k\bigl( z(t) \bigr).
$$

For example,
to sample a flat distribution, $\rho(z)$,
over $[z_c - \Delta z/2, z_c + \Delta z/2]$,
we can use the Legendre polynomials\cite{arfken},
and with $K = 2$,
the bias potential contains only a linear mode for
$R_1(z) = 2 \sqrt{3} \, (z - z_c)/\Delta z$,
which recovers the algorithm by
Langfeld, Lucini and Rago\cite{langfeld2012}.
%
Applied to the energy space, $z = E$,
$c_1(t)/\sigma_z$ represents the scaled local or
statistical temperature\cite{kim2006, *kim2007},
and the inverse-time schedule
was recommended for its update\cite{pellegrini2014}.

Alternatively,
to sample a Gaussian distribution
centered at $z_c$ with width $\sigma_z$
we can use the Hermite polynomials\cite{arfken},
and with $K=3$,
we have a quadratic bias potential
from a linear combination of
$R_1(z) = (z - z_c)/\sigma_z$
and
$R_2(z) = \bigl[(z - z_c)^2 - \sigma_z^2\bigr] /\left(\sqrt 2 \, \sigma_z^2\right)$.
%
This bias potential was commonly empolyed
in umbrella sampling (see e.g. Ref. \cite{zhu2012})
and the method of Gaussian ensemble\cite{hetherington1987,
*challa1988, *costeniuc2006, neuhaus2006, *neuhaus2007},
which is helpful in overcoming barriers
in first-order phase transitions.
%
While the asymptotic updating magntiude
would be given by the inverse-time formula,
we can employ the WL strategy early in simulation
to control the updating magnitude in stages
with the histogram fluctuation redefined by
Eq. \eqref{eq:maxFk}.
%
This method will be referred to as
the adaptive Gaussian ensemble method below.

The use of polynomial basis makes
the bias potential smooth and differentiable,
and hence suitable for MD applications.\cite{zhu2012}
%
However, since we only used two modes
for controlling the mean and variance,
the method only works
for narrow distributions.
%
To sample a wider range,
we can construct an expanded ensemble with
multiple Gaussian distributions
centered at different places via
either the replica exchange
(or parallel tempering)\cite{swendsen1986,
*geyer1991, *hukushima1996, *hansmann1997}
or the simulated tempering\cite{marinari1992,
  *lyubartsev1992, li2007,
  park2007, *nguyen2013, *zhang2015st}
framework.
%
While the mutual exchange probability
can be readily derived in the former
framework\cite{neuhaus2006, *neuhaus2007},
the transition probability in the latter case
requires estimation of the partition function
of the Gaussian ensemble
$Z \equiv \int e^{-\beta \, \mathcal U(\mathbf x)-u(z(\mathbf x),t)} \, d\mathbf x$,
where $\mathbf x$ denotes the molecular coordinates,
$\mathcal U(\mathbf x)$ is the unbiased potential energy function,
$\beta$ is the inverse temperature,
and the collective variable, $z$,
has also been written as a function of $\mathbf x$.
%
The transition probability from
the $i$th Gaussian to the $j$th is given by
is given by
\begin{align*}
A(i\to j) =
\min\left\{1,
  \frac{Z^{(i)}}{Z^{(j)}}
  \exp\left[u^{(i)}\bigl( z(t) \bigr)
   - u^{(j)}\bigl( z(t) \bigr) \right]
  \right\}
  ,
\end{align*}
where $z(t)$ is the current value of $z$.
%
For a neighboring pair, $j = i\pm 1$,
we can estimate the ratio $Z^{(j)}/Z^{(i)}$
from the following approximate formula
%
\begin{align*}
  \ln\frac{ Z^{(j)} }{ Z^{(i)} }
  \approx
  \ln\frac{ \sigma_j }{ \sigma_i }
   -u^{(j)}\bigl( z^{(i,j)} \bigr)
   +u^{(i)}\bigl( z^{(i,j)} \bigr)
  ,
\end{align*}
%
where
$z^{(i,j)} \equiv
\bigl( z_c^{(i)}/\sigma_z^{(i)} + z_c^{(j)}/\sigma_z^{(j)} \bigr) /
\bigl(         1/\sigma_z^{(i)} +         1/\sigma_z^{(j)} \bigr)$.
\note{For implementation, we can further simplify
\begin{align*}
  &-u^{(j)}\bigl( z^{(i,j)} \bigr)
   +u^{(i)}\bigl( z^{(i,j)} \bigr)
   \\
   &=
   (c_1^{(i)} + c_1^{(j)}) \, \Delta z^*
   +
   (c_2^{(i)} - c_2^{(j)}) \, \frac{{\Delta z^*}^2 - 1} {\sqrt 2}
   ,
\end{align*}
where $\Delta z^* = [z_c^{(j)} - z_c^{(i)}]/[\sigma_z^{(j)} + \sigma_z^{(i)}]$.
}

@@

\subsection{\label{sec:lj}
Comparison of updating schemes on a Lennard-Jones fluid}

@@

In Fig. \ref{fig:lj_alpha},
we show the optimal schedules in a few cases.
%
%Generally, the optimal schedule
%for the $\varrho^* = 0.8$ case
%was closer to the inverse-time schedule
%than that for the $\varrho^* = 0.1$ case.
%%
%This was probably because in the former case
%the first mode dominates the total error,
%$\Gamma_1 \approx \Gamma$,
%and the schedule was roughly optimized for this mode,
%resulting in
%$\alpha(t) \approx 1/[\lambda_1 (t + t_0)] \approx 1/(t+t_0)$
%as $\lambda_1 \approx 1$
%[cf. Eq. \eqref{eq:lambda_Gaussian}].
%
Since we have chosen a long equilibration period
for the preliminary run under constant updating magnitude,
the starting magnitude was around
the ideal value of $a_0/2 = 5\times10^{-5}$
given by Eq. \eqref{eq:half_alpha0}.
%
But if the Gaussian were too wide,
e.g. if $\sigma = 0.6$,
the eigenvalues $\lambda_k$ of some modes
became too small to damp out the initial bias,
resulting in much larger initial updating magnitude, $\alpha(0)$
[cf. Eq. \eqref{eq:opt_alpha0}].
%
In this case, the schedule predicted by our method
would be unusable because the resulting deviation in the bias potential
was no longer small,
and one may have to either decrease the width
or to elongate the preliminary simulation under constant magnitude.

@@

\subsection{\label{sec:potts}
Adaptive Gaussian ensemble}

As a proof of principle,
we tested the adaptive Gaussian ensemble method
introduced in Sec. \ref{sec:bandpass}
on the two-dimensional $L\times L$ $10$-state
Potts model\cite{wu1982, newman, wang2001, wang2001pre}
along the energy space.
%
We used $L = 32$ here.
%
We used $26$ Gaussian distributions of width $L$,
and spread them over $[-1.7L^2, -0.9L^2]$.
%
For configuration sampling,
we used the Wolff algorithm\cite{wolff1989, newman}
at temperature $c_1/\sigma_z$
followed by a Metropolis acceptance step
to correct for the quadratic potential, $R_2$.
%
The acceptance ratio for configuration moves
was no less than 90\%.

@@

The relation between error and histogram fluctuation
supports the WL criterion for
checking convergence at stage transitions.
%
For bandpass updating scheme,
we may borrow this criterion
in initial stages, with a modified definition
for the histogram fluctuation
$F^2_{<K}(T) \equiv \sum_{k=1}^{K-1} \tilde F_k^2(T),$
since the histogram flattening
is only partially applied to modes less $K$.
%
One can alternatively define the histogram fluctuation
from the largest mode
%$\max_{k=1, \dots, K-1} \left| \, \tilde F_k(T) \, \right|$.
\begin{align}
  \max_{k=1, \dots, K-1} \left| \, \tilde F_k(T) \, \right|.
  \label{eq:maxFk}
\end{align}
This definition can be readily extended
to the continuous variable case, with
$\tilde F_k(T) = \frac 1 T \sum_{t = 1}^T R_k\bigl( z(t) \bigr)$.
%For a local sampling process,
%we expect that the two definition would behave similarly.
%
\note{
Note that the expected histogram flatness
was previously used to define the error\cite{zhou2005, zhou2008},
and the above demonstration shows the definition
is compatible with ours, Eq. \eqref{eq:error_def}, in the long-time limit.
An interesting application\cite{zhou2008},
of the above result is
to set the updating magnitude by $F/\Gamma$
for some empirical value [$\Gamma \approx 10$ there, see Eq. (8)]
as a substitute of the inverse-time schedule.
}
